Many claims of group polarization are plausibly false due to inappropriate statistics

### 2022-11-24 GroupPolarizationStatmod/paper/ms.tex
some writing up model; figures.R script for making explanatory stat/model figs

Our analysis relies on two important formal concepts: latent opinions and their 
measurement on an ordinal scale. \emph{Latent opinions} are theoretical,
unobservable, psychological, continuous-valued variables that are 
somehow represented and stored in the brain-body, and can be reproduced with
some accuracy when a person is prompted. \emph{Ordinal measurements} of 
of latent opinions are ordinal-valued representations of latent opinions 
in some physical form, such as a mark on a sheet of paper or radio button on a web form, that
indicate the polarity and strength of a person's opinion opinion in terms of
one of a number of opinion bin values.

For example, on a 7-point ordinal scale -3 could indcate 
``strongly disagree'', +3 could be ``strongly agree'', 0 would be psychologically 
neutral, and intermediate values are less extreme opinions of either valence, 
such as was used by \citeA{Moscovici1969}. Or, one might

We assume that participant $i$'s internal, latent psychological opinion at 
time of reporting ($t \in \{pre, post\}$) is drawn from a normal distribution
with mean $\mu_t$ and standard deviation $\sigma_t$,
\begin{equation}
  o_{i,t} \sim \mathcal{N}(\mu_t, \sigma_t).
  \label{eq:opinionDistribution}
\end{equation}
\noindent
This is the condition-level distribution of opinions at time, $t$, chosen to
represent the common experimental design used by group polarization studies,
not because it is statistically appropriate---in fact it is not, which is what
we aim to show in this paper.

@@

>>>>

Our model represents continuous latent opinions and their 
measurement on an ordinal scale. \emph{Latent opinions} are 
unobservable, psychological variables that are 
somehow represented and stored in the brain-body, and can be reproduced with
some accuracy when a person is prompted to give their opinion. 
\emph{Ordinal measurements} of 
of latent opinions are discrete-valued representations of latent opinions 
in some physical form, such as a mark on a sheet of paper or radio button on a web form, that
indicate the polarity and strength of a person's opinion opinion in terms of
one of a number of opinion bin values. All group polarization experiments must somehow
measure opinions, and all but a few use ordinal measurements (those few exceptions include
using bets made during blackjack as a proxy for opinions~\cite{Blascovich1973,Blascovich1975,Blascovich1976}
and using mock jury punitive rewards as a proxy for opinions~\cite{Schkade2000}).

% For example, on a 7-point ordinal scale -3 could indcate 
% ``strongly disagree'', +3 could be ``strongly agree'', 0 would be psychologically 
% neutral, and intermediate values are less extreme opinions of either valence, 
% such as was used by \citeA{Moscovici1969}. Or, one might

We assume that participant $i$ has latent opinion at time $t$ that is drawn from
a group-level normal distribution,
\begin{equation}
  o_{i,t} \sim \mathcal{N}(\mu_t, \sigma_t).
  \label{eq:opinionDistribution}
\end{equation}
\noindent 
This distribution has mean $\mu_t$ and standard deviation $\sigma_t$.
This form is chosen to
reflect the common statistical analysis used by group polarization studies,
not because it is statistically appropriate---in fact it is not, which is what
we aim to show in this paper.

It is appropriate to assume opinions are normally distributed because published
group polarization studies use some sort of $t$-tests, or similar tests, which are designed
to detect whether the mean of two distributions are the same or not. This imposes the
implicit assumption that opinions themselves are normal. But as we will show, 
metric statistical tests such as $t$-tests reliably generate false rejections
of the null hypothesis that there is no difference between two distributions' means.

(HOW SIMPLE CONSENSUS LOOKS LIKE GROUP POLARIZATION)

@@

To evaluate whether a positive group polarization observation is plausibly false,
we develop two layers of evidence for plausible falsehood. The first layer
tests whether we can identify $\sigmapre$ and
